Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2025-01-12 manuscript/manuscript.tex
touch up

\item In addtion to \texttt{incident\_address}, each SR has six
	additional geographic fields:
	\begin{itemize}
		\item{\texttt{latitude} \& \texttt{longitude}}
		\item{\texttt{location}}
		\item{\texttt{street\_name}}
		\item{\texttt{landmark}}
		\item{\texttt{block}}
		\item{\texttt{x\_coordinate\_state\_plane} \& \texttt{y\_coordinate\_state\_plane}}.
 	\end{itemize}

@@

\subsection{Missing Data by Field}
\label{sec:blanks}
Understanding the absence of data by field greatly aids analysis. 
For example, determining if SRs were closed 
before their \texttt{due\_date}, would be challenging as 99.6\% of the
\texttt{due\_date} entries are blank as only DSNY appears to use this field.
Blank (or N/A values) divide into three groups:

\begin{itemize}[left=1.5em]
    \item Mostly Empty: 93-99.9\% blank 
    \item Partially Empty: 40-4\% blank
    \item Few/None Empty: 2-0\% blank
\end{itemize}

\vspace{\baselineskip}
\noindent
The Mostly Empty category includes:
\texttt{taxi\_company\_borough}, 
\texttt{bridge\_highway\_name},
\texttt{due\_date}, and \texttt{road\_ramp}. 
The Partially Empty category includes:
\texttt{cross\_street\_1 \& \_2},
\texttt{location\_type}, and texttt{landmark}. And 
the Few/None Empty category includes \texttt{created\_date}, 
\texttt{complaint\_type},
\texttt{agency}, 
and \texttt{status}. 
Of particular note is that several of the Mostly Empty fields are used by only one or 
two Agencies (See: \autoref{tab:field-usage-summary}).

@@

% Force a blank line and remove indentation
\vspace{\baselineskip}
\noindent
For some analysis efforts, it may be prudent to inquire 
 as to why some fields are mostly/almost always blank. 
Figure~\ref{fig:blank_fields} is a graphic depiction of the \%
empty (blank \& N/As) for each field. It illustrates the grouping into 
Mostly Empty, Partially Empty, and Few/None Empty segments.

@@

\vspace{\baselineskip}
\noindent
The accomanying R program also produces a CSV file showing Field usage by Agency. 
Table~\ref{tab:field-usage-by-agency} contains a sample of Field usage by select Agencies.

\begin{table}[tbp]
\centering
\caption{Sample Field Usage by select Agencies}
\label{tab:field-usage-by-agency}
\small
\begin{tabular}{lrrrrrrr}
\toprule
\textbf{\textbf{Field}} & \textbf{DEP} & \textbf{DHS} & \textbf{DOB} & \textbf{DOT} & \textbf{DPR} & \textbf{DSNY} & \textbf{TLC} \\
\midrule
\texttt{landmark}                  & 2935     & 70185    & 282      & 135150   & 254218   & 424284   & 63133  \\
\texttt{facility\_type}            & 175549   & 0        & 51575    & 159667   & 0        & 66012    & 0      \\
\texttt{due\_date}                 & 0        & 0        & 0        & 0        & 0        & 32097    & 0      \\
\texttt{community\_board}          & 361715   & 99613    & 203715   & 401143   & 266201   & 608188   & 73601  \\
\texttt{vehicle\_type}             & 0        & 0        & 0        & 0        & 0        & 0        & 1696   \\
\texttt{taxi\_company\_borough}    & 0        & 0        & 0        & 0        & 0        & 0        & 4025   \\
\texttt{taxi\_pick\_up\_location}  & 0        & 0        & 0        & 0        & 0        & 0        & 72558  \\
\texttt{bridge\_highway\_name}     & 15       & 21012    & 0        & 7451     & 0        & 613      & 977    \\
\texttt{bridge\_highway\_direction} & 15      & 8973     & 0        & 7450     & 0        & 614      & 1006   \\
\texttt{road\_ramp}                & 15       & 2588     & 0        & 7450     & 0        & 613      & 999    \\
\texttt{bridge\_highway\_segment}  & 15       & 21013    & 0        & 7450     & 0        & 613      & 987    \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Validating Data for Acceptable Values}
\label{sec:domain}
Any analytic effort must ensure that fields containing invalid values 
are identified and isolated from the analysis. For example, the \texttt{latitude} 
and \texttt{longitude} fields were found to all fall within the 
geographic boundaries of New York City. Likewise, 
the \texttt{unique\_key} field was indeed unique, as required. 
Unfortunately, the 311 SR Data Dictionary specifies very few domains
of acceptable values for other fields. Nonetheless, the following
fields were tested and found to comply with their expected
domains, as determined by public
usage and as evidenced in other historical datasets:

\begin{itemize}[left=1.5em]
    \item \texttt{address\_type}
    \item \texttt{status}
    \item \texttt{borough}
    \item \texttt{borough\_boundaries}
    \item \texttt{park\_borough}
    \item \texttt{data\_channel}
    \item \texttt{vehicle\_type}
    \item \texttt{city\_council\_district}
\end{itemize}

\vspace{\baselineskip}
\noindent
However, some fields lacked compliance
to a domain of allowable values, including:

@@

\paragraph{Created and Closed Dates:}
\label{sec:negativeduration}
With the \texttt{created\_date} and \texttt{closed\_date} fields, one 
might expect that these fields would be automatically populated by the  
application software. Thus when creating  or closing an SR 
the the software would automatically populate these fields. 
Unfortunately, this does not seem to be the case, as several anomalies 
exist in the \texttt{created\_date} and \texttt{closed\_date} fields that
an automated process would normally exclude, including:

\begin{itemize}[left=1.5em]
    \item 12,450 SRs with a \texttt{closed\_date} before \texttt{created\_date} creating a nonsensical ``negative duration''. 
    \item Eight SRs with \texttt{created\_date(s)} or \texttt{closed\_date(s)} in 
    the distant past, specifically  \textit{1900-01-01}. Despite few of 
    these entries, such extreme values can greatly distort analytical efforts. 
    \item 163,720 SRs with identical \texttt{created\_date(s)} and \texttt{closed\_date(s)} 
    to the second, creating a nonsensical ``zero duration''. 
    \item 247,000 SRs closed and/or created exactly at midnight, to the second.
\end{itemize}

@@

\paragraph{\texttt{landmark} \& \texttt{street\_name}:} Per the 
311 SR Data Dictionary, the \texttt{landmark} 
field \textit{can refer to any noteworthy location, 
including but not limited to, parks, hospitals, airports, sports facilities, 
performance spaces, etc..} However, most entries are actually 
street names, with a 62\% match to \texttt{street\_name} field. 
Even many of the non-matches (excluding blanks) appear to be 
a logical match, differing only in spelling or formatting 
(e.g., ``NINTH AVE'' vs. ``9 AVE''). This indicates a possible 
misuse of the \texttt{landmark} field as well as a high level of redundancy.

@@

\begin{quote}
\textit{Note: The authors attempted standardization of street names  
in an attempt to match what were obvious identical streets, deemed unequal 
due only to spelling and naming conventions. Those efforts were intended to match, for example, 
\textit{KANE PL} and \textit{KANE PLACE} as well as \textit{AVE P} and \textit{AVENUE P}. However, these 
efforts were  abandoned after they resulted in an increase in street name matches of less than 1\%.)}
\end{quote}

@@

\vspace{\baselineskip}
\noindent
Additionally, certain fields may not be useful for broad analysis
due to their sparse data population. This caveat applies to the fields 
below which are 99\% (or greater) blank.

\vspace{\baselineskip}
\noindent
\begin{quote}
\textit{We should note that even a field that is 99\% 
blank will still contain 64,000 non-blank rows in this 
large (6.4 million row) dataset.; certainly a non-trivia amount of data.}
\end{quote}

\begin{itemize}[left=1.5em]
    \item \texttt{taxi\_company\_borough}
    \item \texttt{road\_ramp}
    \item \texttt{vehicle\_type}
    \item \texttt{due\_date}
    \item \texttt{bridge\_highway\_direction}
    \item \texttt{bridge\_highway\_name}
    \item \texttt{bridge\_highway\_segment}
    \item \texttt{taxi\_pick\_up\_location}
\end{itemize}

>>>>

\item In addtion to \texttt{incident\_address}, each SR has eight
	additional geographic fields:
	\begin{itemize}
		\item{\texttt{latitude} \& \texttt{longitude}}
		\item{\texttt{location}}
		\item{\texttt{street\_name}}
		\item{\texttt{landmark}}
		\item{\texttt{block}}
		\item{\texttt{x\_coordinate\_state\_plane} \& \texttt{y\_coordinate\_state\_plane}}.
 	\end{itemize}

@@

\subsection{Missing Data by Field}
\label{sec:blanks}
Understanding the absence of data by field greatly aids analysis. 
For example, determining if SRs were closed 
before their \texttt{due\_date}, would be challenging as 99.6\% of the
\texttt{due\_date} entries are blank as only DSNY appears to use this field.
Blank (or N/A values) divide into three groups:
\begin{itemize}[left=1.5em]
    \item Mostly Empty: 93-99.9\% blank;
    \item Partially Empty: 40-4\% blank;
    \item Few/None Empty: 2-0\% blank.
\end{itemize}

The Mostly Empty category includes:
\texttt{taxi\_company\_borough}, 
\texttt{bridge\_highway\_name},
\texttt{due\_date}, and \texttt{road\_ramp}. 
The Partially Empty category includes:
\texttt{cross\_street\_1 \& \_2},
\texttt{location\_type}, and texttt{landmark}. And 
the Few/None Empty category includes \texttt{created\_date}, 
\texttt{complaint\_type},
\texttt{agency}, 
and \texttt{status}. 
Of particular note is that several of the Mostly Empty fields are used by only one or 
two Agencies (See: \autoref{tab:field-usage-summary}).

@@

@@

For some analysis efforts, it may be prudent to inquire 
as to why some fields are mostly/almost always blank. 
Figure~\ref{fig:blank_fields} is a graphic depiction of the \%
empty (blank \& N/As) for each field. It illustrates the grouping into 
Mostly Empty, Partially Empty, and Few/None Empty segments.

Table~\ref{tab:field-usage-by-agency} highlights the varied usage of
certain fields across NYC agencies, revealing both specialization and
redundancy. For instance, \texttt{landmark} and
\texttt{facility\_type} are widely used, indicating their importance
in classifying SRs. In contrast, fields like
\texttt{taxi\_company\_borough} and \texttt{vehicle\_type} are used
exclusively by TLC, suggesting that such fields, which serve a narrow
purpose, could be better stored in separate, agency-specific tables.
This approach would reduce the size and complexity of the primary
dataset, making it easier to maintain and analyze while preserving
the ability to perform specialized analyses when necessary.
Streamlining through harmonization and validation would further
enhance consistency and optimize cross-agency analyses.

\begin{table}[tbp]
\centering
\caption{Sample field usage by select agencies among the 6.4 million
  SRs in 2022--2023. }
\label{tab:field-usage-by-agency}
\small
\begin{tabular}{lrrrrrrr}
\toprule
\textbf{\textbf{Field}} & \textbf{DEP} & \textbf{DHS} & \textbf{DOB} & \textbf{DOT} & \textbf{DPR} & \textbf{DSNY} & \textbf{TLC} \\
\midrule
\texttt{landmark}                  & 2935     & 70185    & 282      & 135150   & 254218   & 424284   & 63133  \\
\texttt{facility\_type}            & 175549   & 0        & 51575    & 159667   & 0        & 66012    & 0      \\
\texttt{due\_date}                 & 0        & 0        & 0        & 0        & 0        & 32097    & 0      \\
\texttt{community\_board}          & 361715   & 99613    & 203715   & 401143   & 266201   & 608188   & 73601  \\
\texttt{vehicle\_type}             & 0        & 0        & 0        & 0        & 0        & 0        & 1696   \\
\texttt{taxi\_company\_borough}    & 0        & 0        & 0        & 0        & 0        & 0        & 4025   \\
\texttt{taxi\_pick\_up\_location}  & 0        & 0        & 0        & 0        & 0        & 0        & 72558  \\
\texttt{bridge\_highway\_name}     & 15       & 21012    & 0        & 7451     & 0        & 613      & 977    \\
\texttt{bridge\_highway\_direction} & 15      & 8973     & 0        & 7450     & 0        & 614      & 1006   \\
\texttt{road\_ramp}                & 15       & 2588     & 0        & 7450     & 0        & 613      & 999    \\
\texttt{bridge\_highway\_segment}  & 15       & 21013    & 0        & 7450     & 0        & 613      & 987    \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Validating Data for Acceptable Values}
\label{sec:domain}
Any analytic effort must ensure that fields containing invalid values 
are identified and isolated from the analysis. For example, the \texttt{latitude} 
and \texttt{longitude} fields were found to all fall within the 
geographic boundaries of New York City. Likewise, 
the \texttt{unique\_key} field was indeed unique, as required. 
Unfortunately, the 311 SR Data Dictionary specifies very few domains
of acceptable values for other fields. Nonetheless, the following
fields were tested and found to comply with their expected
domains, as determined by public
usage and as evidenced in other historical datasets:
\begin{itemize}[left=1.5em]
    \item \texttt{address\_type}
    \item \texttt{status}
    \item \texttt{borough}
    \item \texttt{borough\_boundaries}
    \item \texttt{park\_borough}
    \item \texttt{data\_channel}
    \item \texttt{vehicle\_type}
    \item \texttt{city\_council\_district}
\end{itemize}
However, some fields lacked compliance
to a domain of allowable values, as outlined next.

@@

\paragraph{Created and Closed Dates:}
\label{sec:negativeduration}
With the \texttt{created\_date} and \texttt{closed\_date} fields, one 
might expect that these fields would be automatically populated by the  
application software. Thus when creating  or closing an SR 
the the software would automatically populate these fields. 
Unfortunately, this does not seem to be the case, as several anomalies 
exist in the \texttt{created\_date} and \texttt{closed\_date} fields that
an automated process would normally exclude, including:
\begin{itemize}[left=1.5em]
    \item 12,450 SRs with a \texttt{closed\_date} before \texttt{created\_date} creating a nonsensical ``negative duration''. 
    \item Eight SRs with \texttt{created\_date(s)} or \texttt{closed\_date(s)} in 
    the distant past, specifically  \textit{1900-01-01}. Despite few of 
    these entries, such extreme values can greatly distort analytical efforts. 
    \item 163,720 SRs with identical \texttt{created\_date(s)} and \texttt{closed\_date(s)} 
    to the second, creating a nonsensical ``zero duration''. 
    \item 247,000 SRs closed and/or created exactly at midnight, to the second.
\end{itemize}

@@

\paragraph{\texttt{landmark} \& \texttt{street\_name}:} Per the 
311 SR Data Dictionary, the \texttt{landmark} 
field is intended to refer to any noteworthy location, 
including but not limited to parks, hospitals, airports, sports facilities, 
and performance spaces. However, most entries are actually 
street names, with a 62\% match to the \texttt{street\_name} field. 
Many of the non-matches (excluding blanks) also appear to be 
logical matches, differing only in spelling or formatting 
(e.g., ``NINTH AVE'' vs. ``9 AVE''). This indicates a possible 
misuse of the \texttt{landmark} field and highlights significant
redundancy.

@@

We attempted to standardize street names to reconcile entries that
were clearly identical but differed due to spelling or naming
conventions. For example, we sought to match entries such as
\textit{KANE PL} and \textit{KANE PLACE} or \textit{AVE P} and
\textit{AVENUE P}. However, this standardization effort was
ultimately abandoned, as it resulted in less than a 1\% increase in
matched street names.

@@

Additionally, certain fields may not be useful for broad analysis
due to their sparse data population. This applies to the fields 
below which are 99\% (or greater) blank. 
\begin{itemize}[left=1.5em]
    \item \texttt{taxi\_company\_borough}
    \item \texttt{road\_ramp}
    \item \texttt{vehicle\_type}
    \item \texttt{due\_date}
    \item \texttt{bridge\_highway\_direction}
    \item \texttt{bridge\_highway\_name}
    \item \texttt{bridge\_highway\_segment}
    \item \texttt{taxi\_pick\_up\_location}
\end{itemize}
While such sparsity may limit their use in 
broad analyses, these fields can still contain valuable insights 
for specific agency-level studies, given that even a 1\% 
non-blank rate equates to approximately 64,000 rows in this 
6.4 million-row dataset. Proper handling of sparse fields, 
such as storing them in separate tables or using them in 
targeted analyses, could improve data accessibility and 
efficiency without compromising the dataset’s broader utility.
